Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-03-21 doc/invt.tex
small edits to invt.tex

As shown above, with fixed endpoint value of $q(t)$,
the optimization of the error is equivalent to
that of the asymptotic error.
%
We now rewrite Eq. \eqref{eq:error_asym1} as
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt,
  \notag
  %\label{eq:error_asym_action}
\end{equation}
%
where, we have defined the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ).
  \label{eq:error_asym_lagr}
\end{align}
%
Note that the Lagrangian
does not explicitly depend on time, $t$,
so the Hamiltonian,
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \partial \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
is conserved,
and coincides with the Lagrangian.
%
So we may write
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  c^2
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating the square root of the left-hand side, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, t
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is determined by
the boundary condition at $t = T$.
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule, $\alpha(t)$.

@@

We tested the above theory on a one-dimensional system
of $n = 100$ bins,
with a flat target distribution
$p_i = 1/n$.
%
Since in the asymptotic regime,
the distribution converges to $p_i$
no matter the intrinsic distribution, $p^*_i$,
we also set the latter to be flat for convenience.
%
Our goal is to observe how the error
of the bias potential defined in Eq. \eqref{eq:error_sum}
depends on the schedule function $\alpha(t)$
at the end of the simulation.

Initially,
we set the bias potential to zero,
$v_i \equiv 0$,
%
and let the system undergo a period of equilibration
during which the updating magnitude is fixed at $\alpha_0$.
%
After the equilibration,
we reset the origin of the time, $t$,
and started the schedule specified by
Eq. \eqref{eq:alpha_invtlambda}.
%
The goal is then to observe how the final error
depends on the parameter, $\lambda$.
%
Further, we choose $t_0 = 1/(\lambda \, \alpha_0)$
to smooth the transition from equilibration.

Since the error depends on the underlying sampling scheme,
we have applied the above protocol to two
sampling schemes.
%
In both cases, we used the
Metropolis algorithm\cite{metropolis1953, newman, frenkel, landau_binder}
to sample the state space:
%
in each step, a new bin index $j$ was proposed
and then accepted with probability
%
$
A(i \to j) = \min\{ 1, \exp(v_i - v_j) \}.
$
The difference is that,
in the first sampling scheme,
we choose $j$ randomly out of the $n$ bins
to emulate the perfect sampling scheme
discussed in Sec. \ref{sec:Gamma}.
%
We refer to this as the global sampling scheme.
%
In the second scheme,
we limit $j$ to the two neighbors of $i$,
$i \pm 1$, with equal probability $1/2$,
to mimic the one-step sampling scheme
in Sec. \ref{sec:Gamma}.
%
This scheme models local sampling processes
typical in MD and straightforward MC simulations.
%
We refer to this as the local sampling scheme,

@@

Our final observation is that the error functions
in the two cases shared the same shape,
differing only by a multiplicative constant.
%
This again agreed with the theory.
%
Since all eigenvalues of the single-bin scheme were
the same, $\lambda_2 = \cdots = \lambda_n = 1$,
the error in Eq. \eqref{eq:error_invt}
depends on the sampling scheme only through
the sum $\sum_{ k = 2 }^n \Gamma_k$,
which gives rise to the above multiplicative constant.

@@

We verified the optimality of the inverse-time formula
for the Wang-Landau algorithm.

We have shown that
without a priori information,
the single-bin scheme is asymptotically
the best among all updating scheme
in terms of convergence.

For metadynamics, and other multiple-bin updating schemes,
we found formulas to predict the error for finite length
simulations and the optimal relative updating magnitude,
$1/\lambda$.

>>>>

As shown above, with fixed endpoint value of $q(t)$,
the optimization of the error is equivalent to
that of the asymptotic error.
%
We now rewrite Eq. \eqref{eq:error_asym1} as
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt,
  \notag
  %\label{eq:error_asym_action}
\end{equation}
%
where, we have defined the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ).
  \label{eq:error_asym_lagr}
\end{align}
%
Note that the Lagrangian
does not explicitly depend on time, $t$,
so the Hamiltonian,
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \partial \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
is conserved,
and coincides with the Lagrangian.
%
So we may write
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  c^2
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating the square root of the left-hand side, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, t
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is determined by
the boundary condition at $t = T$.
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.

@@

We tested the above theory on a one-dimensional system
of $n = 100$ bins,
with a flat target distribution
$p_i = 1/n$.
%
%Since in the asymptotic regime,
%the distribution converges to $p_i$
%no matter the intrinsic distribution, $p^*_i$,
%we also set the latter to be flat for convenience.
For convenience,
the intrinsic distribution, $p^*_i$,
is a set to be flat as well.
%
Our goal is to observe how the fluctuating error
of the bias potential defined in Eq. \eqref{eq:error_sum}
depends on the schedule, $\alpha(t)$,
at the end of the simulation.

Initially,
we set the bias potential to zero,
$v_i \equiv 0$,
%
and let the system undergo an equilibration period
of constant updating magnitude,
$\alpha(t) = \alpha_0$.
%
After this,
we reset the origin of the time, $t$,
and started the schedule specified by
either
Eq. \eqref{eq:q_opt}
or
Eq. \eqref{eq:alpha_invtlambda}.
%
We use the latter case to verify our theory.
%
We wish to see how the final error
depends on the parameter, $\lambda$,
and if it matches the predicted value.
%
Here, we used $t_0 = 1/(\lambda \, \alpha_0)$
to smooth the transition from equilibration.

Since the error depends on the underlying sampling scheme,
we have applied the above protocol to two
sampling schemes,
a global one and a local one.
%
In both cases, we used the
Metropolis algorithm\cite{metropolis1953, newman, frenkel, landau_binder}
to sample the state space:
%
in each step, a new bin index $j$ was proposed
and then accepted with probability
%
$
A(i \to j) = \min\{ 1, \exp(v_i - v_j) \}.
$
However,
in the first sampling scheme,
we choose the destination $j$
randomly out of the $n$ bins.
%
This sampling scheme,
referred to this as the global sampling scheme,
emulates the perfect sampling scheme
discussed in Sec. \ref{sec:Gamma}
in the long time limit.
%
In the second scheme,
we limit the destination $j$
to the two neighbors of $i$,
$i \pm 1$, with equal probability $1/2$,
to mimic the one-step sampling scheme
in Sec. \ref{sec:Gamma}.
%
This scheme,
referred to this as the local sampling scheme,
models local sampling processes
typical in MD and straightforward MC simulations.
%

@@

Another observation is that the two error functions
differed only by a multiplicative constant.
%
This again agreed with the theory.
%
Since all eigenvalues of the single-bin scheme are
the same, $\lambda_2 = \cdots = \lambda_n = 1$,
the error in Eq. \eqref{eq:error_invt}
depends on the sampling scheme only through
the sum $\sum_{ k = 2 }^n \Gamma_k$,
which gives rise to the above multiplicative constant.

@@

In conclusion,
we have proposed a method of computing
the optimal schedule of the updating magnitude
for a general adaptive FDS simulation.
%
Adaptive FDS methods differ by the updating schemes,
which specify how the bias potential is modified
in each updating step.
%
In the single-bin or WL case,
the update is limited to the current bin;
while in the metadynamics case,
the bias potential of a few neighboring bins to the current one
is updated according to a Gaussian wave packet.

By this method,
we have verified the optimality of
the inverse-time schedule,
Eq. \eqref{eq:alpha_invt},
for the single-bin (WL) case.
%
However,
the optimal schedule,
given by the solution of Eq. \eqref{eq:q_opt},
is algebraically more involved
for a general multiple-bin scheme,
including the Gaussian one used in metadynamics.
%
In the latter case,
both the optimal schedule and the resulting error
are sensitive to the simulation length.

We have also compared different updating schemes.
%
We have shown that in the long time limit
the single-bin scheme belongs to a class of
optimal updating schemes.
%
That is, the single-bin scheme
is asymptotically as efficient as
an equilibrium FDS simulation
with the exact bias potential.
